### 2013-07-25 doc/dissertation/document/chapters/conclusion/conclusion.tex
some cleanup in the concllusion

A second mode of optimization to improve performance regards how
discrete probability distribution functions are sampled in the
implementation. At each transition in the random walk sequence, the
discrete probability distribution function generated by the current
row of the physics operator in which a random walk resides is sampled
to get the next state in the sequence. This operation is also used
when the right hand side of the system is sampled to acquire the
starting state for each random walk in the problem through
Eq~(\ref{eq:adjoint_source_probability}). The current implementation
uses the inverse transform method for sampling where the a random
number is generated and the cumulative distribution function is
queried with a binary seach to find the next discrete state. This
algorithm has $O(log(N))$ time complexity for sampling and $O(N)$ time
complexity for construction. The alias sampling method is suggested
instead as it produces $O(1)$ complexity for sampling and $O(N)$ time
complexity for construction \cite{smith_analysis_2005}. Profiling of
the implementation shows that this improvement could also provide
quality reductions in run time.

@@

All methods required quality preconditioning. Even a Krylov method may
be effectively useful without it. However, these methods typically
converge eventually whereas quality preconditioning may not be enough
to gaurantee convergence of MCSA. We saw in our work that in many
cases very agressive preconditioning was not enough to condition the
eigenvalues of the system into a regime where MCSA would apply. The
limitation that the spectral radius of the physics operator iteration
matrix must be less than one is by far the most prohibitive component
of MCSA. The fact that the Monte Carlo sequence is bsed on the
Richardson iteration means that without sufficient preconditioning,
even if this condition is achieved convergence may be incredibly
slow. Therefore, we propose two elements of future work that aim to
move away from this restriction, both fundamentally based on sampling
a sequence not based on the Neumann series.

@@

Iterative methods of the $n^{th}$ degree produce solutions at each
iteration of the following form:
\begin{equation}
  \ve{x}^{k+1} = F(\ve{A},\ve{b},\ve{x}^{k},\ve{x}^{k-1},\dots,\ve{x}^{k-n+1})\:,
\label{eq:nth_degree_iter}
\end{equation}
where $\ve{A}$ is the linear operator $\ve{b}$ is the right hand side
as defined by Eq~(\ref{eq:linear_problem}). Richardson's iteration is
an iterative method of the first degree such that $\ve{x}^{k+1} =
F(\ve{A},\ve{b},\ve{x}^{k})$. To move the Monte Carlo method away from
the spectral radius limitation and therefore we seek higher order
methods as outlined by Hallett \cite{hughes_hallett_second-order_1984}
that have garanteed convergence at all eigenvalue spectra. In his
work, Hallett builds a second order method such that $\ve{x}^{k+1} =
F(\ve{A},\ve{b},\ve{x}^{k-1})$. Such a method results from mapping the
original Richardson iteration into a different of eigenvalues in the
complex plane that have a spectral radius less than one, thus
garanteeing convergence. Derived as expansions of the iteration
sequence in Chebyshev polynomials, Hallett arrives at simple
formulation for the second degree method that requires knowledge of
the bounding curve in the complex plane of the eigenvalue spectra for
good convergence properties.

As it turns out, Dimov has already applied such a technique to the
Monte Carlo method but arrived at the same conclusions in a different
way \cite{dimov_parallel_2001,dimov_new_1998}. In his work, Dimov
instead formulated the random walk sequence by instead building the
Neumann series from the \textit{resolvent} of the operator. The
resolvent operator is defined as that which spans all of the complex
plane but has singularities at the points that are eigenvalues of the
original operator. By building the Monte Carlo sequence from this
operator, Dimov arrived at a Monte Carlo method which in fact was
equivalent to a fixed point method of the second degree and utilized
effectively the same spectral mapping procedure as Hallett. Just like
the deterministic case, some knowledge of the eigenvalue spectrum of
the system was required to ensure convergence. These methods should be
researched and developed further in the context of MCSA as a potential
means of always gauranteeing convergence of the algorithm regardless
of the eigenvalue spectrum.

@@

As a final means of improvement we consider stochastic projection
methods as developed by Sabelfeld \cite{hutchison_stochastic_2011}. In
his work, Sabelfeld claims his method is 'beyond Markov chains and von
Neumann-Ulam scheme'. To a certain extent this is true, his method is
based on a projection method and should always converge. In his work,
he utilizes a Kaczmarz iteration:
\begin{equation}
  \ve{x}^{k+1} = \ve{x}^k + \lambda_k \frac{b_i - \langle \ve{a}_i
    \ve{x}^k \rangle}{||\ve{a}_i||}\ve{a}_i\:,
  \label{eq:kaczmarz_iteration}
\end{equation}
where $\ve{a}_i$ is the $i^{th}$ row vector of the operator and
$\lambda_k$ is the relaxation parameter.

The stochastic nature of the algorithm comes from effectively
modifying the algorithm such that the value of $i$ is now selected at
random. This is interesting but arguably does not serve our purposes
at well. We have an interest in a fully stochastic process like a
random walk sequence because of its potential benefits in terms of
resiliency and the ability to use superposition to combine results
from replicated problems to enhance performance. In Sabelfeld's work,
the randomness is used to select which deterministic iteration step
will be performed and does not meet either of these
requirements. However, that does not mean there is nothing to take
from this work. If one writes Eq~\ref{eq:kaczmarz_iteration} in terms
of applications of projection operators as that outlined in
\cite{tanabe_projection_1971} we see that this actually takes a
similar form as the Neumann series and could therefore be used in a
random walk sequence. Research studying random walks based on
projection sequences is suggested as a means of moving beyond the
spectral radius limitation.

@@

Before this work was performed, MCSA had yet to be applied to the
neutron transport problem, asymmetric systems, and nonlinear problems
including fluid flow systems. Techniques were developed by this work
that enabled solutions for all of these problems using MCSA. In many
cases, it was found that the iterative performance of the MCSA-based
methods were superior to that of conventional solution techniques
based on Krylov methods. Furthermore, with improved implementation
performance, future problems may be solved in less time and more
efficiently using MCSA due to this demonstrated improved iterative
performance.

When this work began, the Neumann-Ulam sequence and MCSA in general
had yet to be parallelized in a domain decomposed framework. This
research developed a new parallel algorithm based on Monte Carlo
particle transport that permitted MCSA solutions to be generated on
$O(100,000)$ cores on a leadership class computing facility. The new
algorithm was also demonstrated to achieve better strong and weak
scaling than conventional Krylov methods at high levels of
concurrency. If future work can improve the time to solution using
MCSA through optimization, then this improved scaling demonstrated by
the algorithm can potentially improve the time to solution for the
engineer and more efficiently leverage high performance computing
hardware.

>>>>

A second mode of optimization to improve performance regards how
discrete probability distribution functions are sampled in the
implementation. At each transition in the random walk sequence, the
discrete probability distribution function generated by the current
row of the physics operator in which a random walk resides is sampled
to get the next state in the sequence. This operation is also used
when the right hand side of the system is sampled to acquire the
starting state for each random walk in the problem through
Eq~(\ref{eq:adjoint_source_probability}). The current implementation
uses the inverse transform method for sampling where the a random
number is generated and the cumulative distribution function is
queried with a binary seach to find the next discrete state. This
algorithm has $O(log(N))$ time complexity for sampling and $O(N)$ time
complexity for construction. The alias sampling method is suggested
instead as it produces $O(1)$ complexity for sampling and $O(N)$ time
complexity for construction \cite{smith_analysis_2005}. Sabelfeld
outlines the use of this sampling technique in his Monte Carlo
algorithms \cite{sabelfeld_sparsified_2009}. Profiling of the
implementation shows that this improvement could also provide quality
reductions in run time.

@@

All numerical methods for solving linear problems require quality
preconditioning. Even a Krylov method may be effectively useful
without it. However, these methods typically converge eventually
whereas quality preconditioning may not be enough to gaurantee
convergence of MCSA. We saw in our work that in many cases very
agressive preconditioning was not enough to condition the eigenvalues
of the system into a regime where MCSA would apply. The limitation
that the spectral radius of the physics operator iteration matrix must
be less than one is by far the most prohibitive component of MCSA. The
fact that the Monte Carlo sequence is based on the Richardson
iteration means that without sufficient preconditioning, even if this
condition is achieved convergence may be incredibly slow. Therefore,
we propose two elements of future work that aim to move away from this
restriction, both fundamentally based on sampling a sequence not based
on the Neumann series.

@@

Iterative methods of the $n^{th}$ degree produce solutions at each
iteration of the following form:
\begin{equation}
  \ve{x}^{k+1} = F(\ve{A},\ve{b},\ve{x}^{k},\ve{x}^{k-1},\dots,\ve{x}^{k-n+1})\:,
\label{eq:nth_degree_iter}
\end{equation}
where $\ve{A}$ is the linear operator $\ve{b}$ is the right hand side
as defined by Eq~(\ref{eq:linear_problem}). Richardson's iteration is
an iterative method of the first degree such that $\ve{x}^{k+1} =
F(\ve{A},\ve{b},\ve{x}^{k})$. To move the Monte Carlo method away from
the spectral radius limitation and therefore we seek higher order
methods as outlined by Hallett \cite{hughes_hallett_second-order_1984}
that have gauranteed convergence at all eigenvalue spectra. In his
work, Hallett builds a second order method such that $\ve{x}^{k+1} =
F(\ve{A},\ve{b},\ve{x}^{k-1})$. Such a method results from mapping the
original Richardson iteration into a different of eigenvalues in the
complex plane that have a spectral radius less than one, thus
garanteeing convergence. Derived as expansions of the iteration
sequence in Chebyshev polynomials, Hallett arrives at simple
formulation for the second degree method that requires knowledge of
the bounding curve in the complex plane of the eigenvalue spectra for
good convergence properties.

As it turns out, Sabelfeld has already applied such a technique to the
Monte Carlo method but arrived at the same conclusions in a different
way \cite{sabelfeld_sparsified_2009}\footnote{Although the technique
  is outlined in this document, Sabelfeld cites his previous work in
  this area in 1982.} with this work effectively re-developed by Dimov
without citation of Sabelfeld
\cite{dimov_parallel_2001,dimov_new_1998}. In his work, Sabelfeld
formulated the random walk sequence by instead building the Neumann
series from the \textit{resolvent} of the operator. The resolvent
operator is defined as that which spans all of the complex plane but
has singularities at the points that are eigenvalues of the original
operator. By building the Monte Carlo sequence from this operator,
Sabelfeld arrived at a Monte Carlo method which in fact was equivalent
to a fixed point method of the second degree and utilized effectively
the same spectral mapping procedure as Hallett. Just like the
deterministic case, some knowledge of the eigenvalue spectrum of the
system was required to ensure convergence. These methods should be
researched and developed further in the context of MCSA as a potential
means of always gauranteeing convergence of the algorithm regardless
of the eigenvalue spectrum.

@@

As a final means of improvement we consider stochastic projection
methods as developed by Sabelfeld \cite{hutchison_stochastic_2011}. In
his work, Sabelfeld claims his method is 'beyond Markov chains and von
Neumann-Ulam scheme'. To a certain extent this is true, his method is
based on a projection method and should always converge. In his work,
he utilizes a Kaczmarz iteration:
\begin{equation}
  \ve{x}^{k+1} = \ve{x}^k + \lambda_k \frac{b_i - \langle \ve{a}_i
    \ve{x}^k \rangle}{||\ve{a}_i||}\ve{a}_i\:,
  \label{eq:kaczmarz_iteration}
\end{equation}
where $\ve{a}_i$ is the $i^{th}$ row vector of the operator and
$\lambda_k$ is a relaxation parameter.

The stochastic nature of the algorithm comes from effectively
modifying the algorithm such that the value of $i$ is now selected at
random. This is interesting but arguably does not serve our purposes
at well. We have an interest in a fully stochastic process like a
random walk sequence because of its potential benefits in terms of
resiliency and the ability to use superposition to combine results
from replicated problems to enhance performance. In Sabelfeld's work,
the randomness is used to select which deterministic iteration step
will be performed and does not meet either of these
requirements. However, that does not mean there is nothing to learn
from this work. If one writes Eq~\ref{eq:kaczmarz_iteration} in terms
of applications of projection operators as that outlined in
\cite{tanabe_projection_1971} we see that this actually takes a
similar form as the Neumann series and could therefore be used in a
random walk sequence. In addition, perhaps the above algorithm can be
modified such that the randomness of the projection sequence permits
improved results via the superposition principle. Research studying
random walks based on projection sequences is therefore suggested as a
means of moving beyond the spectral radius limitation.

@@

Before this work was performed, MCSA had yet to be applied to the
neutron transport problem, asymmetric systems, and nonlinear problems
including fluid flow systems. Techniques was developed by this work
that enabled solutions for all of these problems using MCSA. In many
cases, it was found that the iterative performance of the MCSA-based
methods were superior to that of conventional solution techniques
based on Krylov methods. Furthermore, with improved implementation
performance, future problems may be solved in less time and more
efficiently using MCSA due to this demonstrated improved iterative
performance.

When this work began, the Neumann-Ulam sequence and MCSA in general
had yet to be parallelized in a domain decomposed framework. This
research developed a new parallel algorithm based on Monte Carlo
particle transport that permitted MCSA solutions to be generated on
$O(100,000)$ cores on a leadership class computing facility. The new
algorithm was also demonstrated to achieve better strong and weak
scaling than conventional Krylov methods at high levels of
concurrency. If future work can improve the time to solution using
MCSA through optimization, then this improved scaling demonstrated by
the algorithm can potentially improve the time to solution for the
engineer and more efficiently leverage high performance computing
hardware for analysis.
